Dialect: A Compact Line Notation Based on SMILES

### 2022-03-29 manuscript.tex
Improves the flow of Implict Hydrogen Count.

To qualify for implicit hydrogen counting, an atom must have a \ttt{hydrogens} attribute set to \ttt{Implicit}. As noted previously, this setting requires: (1) \ttt{isotope}, \ttt{charge}, and \ttt{configuration} must be set to their default values; and (2) the value of \ttt{element} must equal one of: \ttt{B}; \ttt{C}; \ttt{N}; \ttt{O}; \ttt{P}; \ttt{S}; \ttt{F}; \ttt{Cl}; \ttt{Br}; \ttt{I}; or \ttt{None}.

For unselected atoms, implicit hydrogen count equals subvalence, with one exception. An oxygen atom with one single bond has an implicit hydrogen count of one because subvalence is one ($2 - 1$). However, an oxygen atom with two singly-bonded neighbors has an implicit hydrogen count of zero ($2 - 2$). For atoms whose \ttt{element} attribute is \ttt{None}, implicit hydrogen count equals zero.

For selected atoms, a modified procedure is used. If subvalence is greater than zero, then the implicit hydrogen count equals subvalence minus one. This subtraction models the unpaired electron that would would be required for delocalized bonding, or localized bonding if the atom were deselected. Consider a selected carbon atom in benzene with a \ttt{hydrogens} attribute of \ttt{Implicit}. Subvalence equals two, so implicit hydrogen count equals one ($2 - 1$).

If the subvalence of a selected atom equals zero, then an implicit hydrogen count of zero is reported. The meaning of such an atomic state may seem suspect. If an atom has no unpaired electrons, how can it participate in delocalized bonding? As will be explained in detail later (Pruning), this situation can arise for reasons of convenience or tradition. Returning zero avoids miscalculation of the implicit hydrogen count.

An implicit hydrogen count may or may not correlate with chemical intuition or experimental data. Consider the phosphorous-bearing atom of hypophosphorous acid (HOP(O)H\textsubscript{2}). We might expect the implicit hydrogen count to equal the experimentally-determined hydrogen count (2). However, the subvalence for the phosphorous atom is found to be three ($2 + 1 - 3$). The implicit hydrogen count is therefore zero ($3 - 3$) rather than the expected two. The hydrogen count of this atom, and others like it, must be encoded as a virtual hydrogen count.

>>>>

An atom whose \ttt{hydrogens} attribute equals \ttt{Implicit} signals that its hydrogens must be counted algorithmically. The exact algorithm depends on the value of the atoms's \ttt{selected} attribute.

For an unselected atom, implicit hydrogen count equals subvalence. For example, an oxygen atom with one single bond has an implicit hydrogen count of one because subvalence is one ($2 - 1$). Similarly, an oxygen atom with two singly-bonded neighbors has an implicit hydrogen count of zero ($2 - 2$).

For a selected atom, implicit hydrogen count equals subvalence minus one. This subtraction accounts for the extra valence implied by the atom's membership in the delocalization subgraph. Consider a selected carbon atom in benzene with a \ttt{hydrogens} attribute of \ttt{Implicit}. Subvalence equals two ($4 - 2$), so implicit hydrogen count equals one ($2 - 1$). Likewise, the subvalence of a selected nitrogen atom in pyridine equals one ($3 - 2$) so implicit hydrogen count equals zero ($1 - 1$). Stated differently, the calculation ensures that an atom's implicit hydrogen count is identical before and after deselection.

If the subvalence of a selected atom equals zero, then an implicit hydrogen count of zero is reported. The meaning of such an atomic state may seem suspect because an atom without a free valence can not perform the required promotion of a single bond during deselection. As will be explained in detail later (Pruning), this situation can arise for reasons of convenience or tradition. Returning zero avoids miscalculation of the implicit hydrogen count.

An atom's implicit hydrogen count may or may not correlate with chemical intuition or experimental data. Consider the phosphorous-bearing atom of hypophosphorous acid (HOP(O)H\textsubscript{2}). We might expect the implicit hydrogen count to equal the experimentally-determined hydrogen count (2). However, the subvalence for the phosphorous atom is found to be three ($2 + 1 - 3$). The implicit hydrogen count is therefore zero ($3 - 3$) rather than the expected two. To represent this atom and other like it, virtual hydrogens must be used.
